PYG4OMETRY: a PYTHON library for the creation of Monte Carlo radiation transport physical geometries

### 2020-09-03 public/paper/pyg4ometry.tex
Paper update

@@

\title{Pyg4ometry: a Python library for the creation of Monte Carlo radiation transport physical geometries}

@@

% \linenumbers

@@

\begin{small}
\noindent
{\em Program Title: pyg4ometry }                                         		\\
{\em Licensing provisions: GPLv3 }			\\
{\em Programming language: python, C/C++}                         		\\

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, radiation transport (RT) programmes, these 
include MCNP \cite{Mcnp_Werner}, FLUKA \cite{Fluka_Ferrari,Fluka_Bohlen}, Geant3 \cite{Geant3_Brun} and Geant4 \cite{Geant4_Agostinelli}. 
All these codes are based on the Monte Carlo technique but each code either has a particular specialism, simulation methodology or target user community.  
RT Monte Carlo (RTMC) simulations have diverse uses, shielding calculations for radiological protection, detector performance, medical imaging and therapy and space radiation environment  simulation are some examples. A fundamental requirement of all of the codes is to supply a computer readable description of the 
physical geometry that  the particles are passing through.  The creation of geometry files is typically a very time consuming activity and the simulation validity 
and performance is directly dependent on the quality of the geometry. The requirement of a geometry system that allows the conversion between files prepared 
for different codes will enable cross checks of the physics processes in different particle transport codes. Parallel to the creation of geometry files for RT 
programs, usually computer assisted design (CAD) files exist for systems which need to be simulated in RT programmes. The fundamental geometric 
representations in CAD files is typically not amenable for RT programmes.  For these reasons it is advantageous to create a software tool that allows 
particle transport code users to rapidly develop geometry files, convert between common RT programme geometry formats and finally import CAD models. 
\colorbox{yellow}{More explanation of Geant4 and FLUKA}

This paper describes a geometry creation and conversion package called {\em pyg4ometry}, implemented in {\em Python} and internally based on the Geant4 geometry GDML \cite{GDML}. The main features of  {\em pyg4ometry} are a python scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry descriptions, conversion from CAD formats  (STEP and IGES) based on FreeCAD \cite{FreeCAD} and OpenCascade \cite{OpenCASCADE} and powerful geometry visualisation tools based on VTK \cite{VTK4}. The origin of pyg4ometry was a set of utilities to prepare geometry for an accelerator beam line simulation based on Geant4 called BDSIM \cite{BDSIM_Nevay}. Accelerator physicists like specialists in other areas need a tool to quickly simulate beam losses, and the subsequent interaction of the charged particle beam with the accelerator material. Figure~\ref{fig:workflow} depicts various workflows possible with {\em pyg4ometry}. {\em Pyg4ometry} is a not a software package but a toolkit library for creation and conversion of geometry, a user would typically write very small python programs to use the {\em pyg4ometry} toolbox. This paper describes version 1 of pyg4ometry, which is freely available as a git prepository and via pypi.

\begin{figure*}[hbt!]
  \normalsize
  \centering
  \includegraphics[width=0.7\textwidth]{./diagrams/workflow.pdf}
  \caption{\label{fig:workflow}Pyg4ometry geometry conversion workflow.}
\end{figure*}

There are existing codes that have functionality similar to pyg4ometry. Tools exist to convert CAD files to GDML, open source examples are GUIMesh \cite{GUIMesh_Pinto}, commercial products include ESABASE2 \cite{ESABASE2} and FASTRAD \cite{FASTRAD}. There are also tools from the fusion and 
neutronics community which  can convert CAD geometry into formats usable by RTMC codes, examples include DAGMC \cite{DAGMC} and McCad \cite{McCad}.  
CAD software can export data to STL. If a CAD model comprises of a large number of parts, exporting, assigning material to and placing the STL  components into 
the RTMC can be a cumbersome task. The existent set of software does not provide a complete set of tools to efficiently create complex geometries.

This paper is structured as follows, first a brief radiation transport-centric introduction to computer descriptions of geometry, followed by an explanation of the design and implementation of {\it Pyg4ometry}. Thesections that follow describe how {\it pyg4ometry} can be used to perform rapid geometric modelling, conversion from FLUKA to GDML, GDML to FLUKA and CAD to GDML. The paper concludes with an example of composite complex system consisting components of all the possible geometry input files formats.

\section{Computer descriptions of geometry} \label{sec:geometric}
Central to a computer readable geometry of a system is how a solid is defined in three dimensions. There are numerous different ways to describe a
geometry, including constructive solid geometry, boundary representation and tessellated polygons, these are described briefly in the following sections.  
The Geant4 geometry specification is a mixture of geometry modelling techniques and described in detail last.

%%%%%%%%%%%%%%%%% Constructive solid geometry (CSG) 
Constructive solid geometry uses boolean operations, subtraction, intersection and union, between simple solid shapes (e.g. cube, cylinder, sphere etc) or infinite
volumes (e.g a plane defined infinite half-space) to model complex surfaces which represent a solid. Boolean operations and solids can be combined to form a 
CSG-tree to model complex geometry. FLUKA uses CSG to model solids, the form of boolean expression used by FLUKA is not a general CSG-tree but a 
logical expression in disjunctive normal-form.

@@

%%%%%%%%%%%%%%%%% Tessellated polygons
Solid volumes can be defined using triangular, quadrilateral or tetrahedral meshes. Numerous formats exist to describe meshes, the ubiquitous being STL and 
more modern variants being PLY, OBJ, etc. For solids with curved faces a tessellated mesh will always give an approximate description. As the mesh 
deviation distance from the solid decreases the number of polygons increases and hence the memory consumption and execution time of the RTMC.

%%%%%%%%%%%%%%%%% Geant4 geometery
Geant4 geometry description is a mixture of BREP, CSG and tessellated concepts. Geant4 includes XX fundamental solids, although it does not store a sense 
of topology present in traditional CAD BREP systems. One of the fundamental solids is a {\em TessellatedSolid} which can used to represent STL or PLY files. 
Geant4 also provides the ability to perform boolean operations on these primitive solids. The richest and most flexible geometry description is currently used by
Geant4. Not only do solid objects need to be defined but also placed in a world coordinate system. Geant4 has two concepts which facilitate this {\it Logical Volumes} and {\it Physical Volumes}. A Logical Volume is a region of space  that is defined by an outer solid but also other attributes like material, magnetic field 
and zero or more physical {\it daughter} volumes. A Physical Volume is a placement (or instance) of a Logical Volume.

%%%%%%%%%%%%%%%%% Geant Detector Mark-up Language
To exchange geometry descriptions Geant Detector Mark-up Language (GDML) was developed \cite{GDML}. GDML is an XML based description of 
Geant4 geometry description. Geant4 and ROOT \cite{fons_rademakers_2019_3895860} can read and write GDML format and it is commonly used as an exchange format for Geant4 geometries.

\section{Pyg4ometry design and layout}
Pyg4ometry is a python package consisting of semi-independent modules. The module \verb|pyg4ometry.geant4| contains all classes for geant4 detector 
construction, \verb|pyg4ometry.gdml| for GDML reading and writing. Modules for importing and exporting other geometry formats  \verb|pyg4ometry.fluka|, 
\verb|pyg4ometry.stl| and \verb|pyg4ometery.freecad|.

The core of {\em pyg4ometry} consists of python classes that mimic Geant4 solids, logical, physical volume objects, GDML parameters and material classes.
The constructors of the python classes are kept as close to the original Geant4 C++ implementation so that pyg4ometry users do not have to learn 
a new API. For example the \verb|G4Box| class in Geant4, has the XML tag \verb|Box| in GDML and is represented by the \verb|geant4.solid.Box| class in 
Python. Geometry construction in python proceeds in a way which is very similar to geometry construction in Geant4. A user relatively familiar with Geant4
should be able to start creating geometry in pyg4ometry immediately. In the following sections novel or important developments in pyg4ometry are described.

%\begin{table}[hbt!]
%\centering
%\begin{tabular}{ l l l } \hline
%GDML tag 		& Child tags				& pyg4ometry class 				\\ \hline
%Define 			& Constant				& gdml.Constant				\\
%				& Variable					& gdml.Variable					\\
%				& Quantity					& gdml.Quantity				\\
%				& Expression 				& gdml.Expression				\\
%				& Position					& gdml.Position					\\
%				& Rotation					& gdml.Rotation				\\
%				& Scale					& gdml.Scale					\\
%				& Matrix					& gdml.Matrix					\\ \hline
%Materials 		& Element					& geant4.Element				\\
%				& Isotope					& geant4.Isotope				\\
%				& Material					& geant4.Material				\\ \hline
%Solid  			& Box					& geant4.solid.Box				\\
%	  			& Tube					& geant4.solid.Tubs				\\
%	  			& CutTube				& geant4.solid.CutTubs			\\
%	  			& Cone					& geant4.solid.Cons				\\
%	  			& Para					& geant4.solid.Para				\\
%	  			& Trd					& geant4.solid.Trd				\\
%	  			& Trap					& geant4.solid.Trap				\\
%	  			& Sphere					& geant4.solid.Sphere			\\
%	  			& Orb					& geant4.solid.Orb				\\
%	  			& Torus					& geant4.solid.Torus				\\
%	  			& Polycone				& geant4.solid.Polycone			\\
%	  			& GenericPolycone			& geant4.solid.GenericPolycone	\\
%	  			& Polyhedra				& geant4.solid.Polyhedra			\\
%				& GenericPolyhedra                  & geant4.solid.GenericPolyhedra	\\ 
%	  			& Eltube					& geant4.solid.EllipticalTube		\\
%	  			& Ellipsoid					& geant4.solid.Ellipsoid			\\
%	  			& Elcone					& geant4.solid.EllipticalCone		\\
%				& Paraboloid				& geant4.solid.Paraboloid			\\
%				& Hype					& geant4.solid.Hyperboloid		\\
%				& Tet						& geant4.solid.Tet				\\
%				& Xtru					& geant4.solid.ExtrudedSolid		\\
%				& TwistedBox				& geant4.solid.TwistedBox		\\
%				& TwistedTtap				& geant4.solid.TwistedTrap		\\
%				& TwistedTrd				& geant4.solid.TwistedTrd 		\\
%				& TwistedTube				& geant4.solid.TwistedTube		\\
%				& Arb8					& geant4.solid.GenericTrap		\\
%				& Tessellated				& geant4.solid.TessellatedSolid 	\\
%				& Union					& geant4.solid.Union				\\
%				& Subtraction				& geant4.solid.Subtraction			\\
%				& Intersection				& geant4.solid.Intersection 		\\
%				& MultiUnion				& geant4.solid.MultiUnion 			\\
%				& ScaledSolid				& geant4.solid.ScaledSolid		\\				
%	  			& OpticalSurface			& geant4.solid.OpticalSurface		\\ \hline
%Structure		& Volume					& geant4.LogicalVolume			\\
%				& Assembly				& geant4.AssemblyVolume 		\\
%				& PhysVol					& geant4.PhysicalVolume			\\ 
%				& ReplicaVol				& geant4.ReplicaVolume			\\ 
%				& ParaVol					& geant4.ParameterisedVolume	\\ 
%				& DivisionVol				& geant4.DivisionVolume			\\
%				& SkinSurface				& geant4.SkinSurface			\\ 
%				& BorderSurface			& geant4.BorderSurface			\\ \hline
%
%\end{tabular}
%\label{tab:gdml-tags}
%\caption{GDML tags and corresponding pyg4ometry classes.}
%\end{table}

%%%%%%%%%%%%%%%%%  Input and output  
For each input format available in pyg4ometry a dedicated {\it Reader} class is implemented, so \verb|gdml.Reader|, \verb|stl.Reader|, \verb|fluka.Reader| 
and \verb|freecad.Reader|. Each reader constructs the appropriate Geant4 classes and provides a \verb|Registry| which can be used or manipulated by the user. 
Output consists of taking the registry and writing the appropriate GDML.

%%%%%%%%%%%%%%%%%  Internal data representation
The internal data representation follows closely the structure of GDML. A \verb|Registry| class aggregates python ordered dictionaries that are  used to store the main 
elements of a GDML file. As a user specifies geometry the registry is updated. When a user is finished with the geometry, the registry is written to disk as a GDML file.
\verb|Registry| instances can be aggregated to form a composite geometry or volumes can be removed and added.

@@

\begin{lstlisting}[caption={A simple python script using pyg4ometry to create a simple Geant4 geometry.},label={lst:gdmlExpressions}, language=Python]
# Import modules 
import pyg4ometry

@@

%%%%%%%%%%%%%%%%%  Replica, division, parametrised and looped volumes
A powerful feature of Geant4 and hence GDML is the ability to either repeat, divide or parametrise geometry. The create multiple replicas of a volume in a cartesian, cylindrical or spherical grid is known as a ReplicaVolume. A DivisionVolume breaks a primitive into segments in either a cartesian or cylindrical polar coordinates. 
A Parametrised Volume allows for the arbitrary multiple placement of solids where the parameters are allowed to vary for each placement.  GDML supports loops, 
where sections of GDML are allowed to be repeated with varying parameters. GDML loops are not supported by pyg4ometry and will be implemented in a 
future release.

\subsection{Tessellation of solids (meshing)}
For each Geant4 primitive solid a triangular tessellated Vertex-Face mesh is generated and cached. This mesh is then used to determine the extent of placed 
instances of geometry (physical volumes) and meshes for CSG derived solids. CSG mesh calculations are performed using a BSP tree technique in Python 
\cite{pycsg} or the Computational Geometry Algorithms Library (CGAL) surface meshes \cite{cgal:bsmf-sm-20b} . Triangular meshes based on CSG 
operations involving curved surfaces often contain large numbers of thin triangles. Before meshes are visualised or written to disk various polygon mesh 
algorithms  from CGAL \cite{cgal:lty-pmp-20b} can be employed to give the meshes more desirable features.

@@

The python version of Geant4 physical volume is also used to store visualisation attributes like the solid's 
colour, surface or wireframe representation and visibility. Overlaps detected in the mesh geometry are
stored in the Logical Volume and can be  displayed separately to allow a user to understand the reason for 
the overlaps.

%%%%%%%%%%%%%%%%% Adding properties to volumes
Geometry alone is an insufficient information for a RTMC simulation. Often other attributes need to be assigned to regions of space for example material definition, magnetic field or optical properties. These physical properties can be used to define the visualisation attributes of a volume.

\subsection{Overlap detection}
All RTMC transport codes can not handle spatial overlap between two geometric objects and likely are to have ill defined behaviour when tracking particles  
through such a situation.  A key feature of pyg4ometry is the detection of potential overlaps efficiently, it does this by performing an intersection operation etween solid instances and determining if the resulting mesh is empty. Figure~\ref{fig:overlap} shows the potential overlaps, a) protrusion of a daughter from the mother, b) and intersection with finite volume and c) an overlap where two daughters share a face. 
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{./diagrams/overlap.pdf}
\caption{Schematic of the three different types of overlap between daughters of a mother logical volume.}
\label{fig:overlap}
\end{center}
\end{figure}

@@

Overlap detection in pyg4ometry relies of the meshes generated for each solid, 
any detection algorithm will be approximate. Generally the overlap detection 
algorithm proceeds as 
\begin{enumerate}
\item Check if any of the daughter solid meshes have a non-null intersection with the compliment of mother solid mesh
\item Loop over all daughter volumes and check
\item Check
\end{enumerate}

@@

%%%%%%%%%%%%%%%%% Graphical user interface
% A basic Qt based graphical user interface is also available with pyg4ometry. If a model has a large number of volumes defined in a hierarchical tree structure, it is % cumbersome for the user to find the logical or physical volume.

\section{Rapid geometry modelling}
Given the python scripting interface, expression  and tessellation engines. It is possible for a user to rapidly specify the geometrical layout of the RT problem, vary 
the parameters of the geometry and visualise the geometry. Finally when the desired result without geometry overlaps is finalised a GDML file can be output
from the internal memory representation. Some example code is presented in Listing~\ref{lst:pythonRapidModelling}. The structure should be familiar to regular 
users of Geant4, first a \verb|Registry| need to be created to store all the pyg4ometry objects; followed by constants;  then materials and properties, solids and logical and physical volumes; finally the whole geomtery can be saved as a GDML file and/or visualised using VTK.

\begin{lstlisting}[caption={A simple python script using pyg4ometry to create a simple Geant4 geometry.},label={lst:pythonRapidModelling}, language=Python]
# Import modules 
import pyg4ometry.gdml as gd
import pyg4ometry.geant4 as g4
import pyg4ometry.visualisation as vi

@@

The pyg4ometry python code in the example is approximately as expressive 
as the GDML it writes. The benefit of wrapping GDML/Geant4 in Python allows 
to very rapid prototyping of geometry without concerns of compilation of C++ (in the 
case of Geant4) or writing well formed XML (in the case of GDML). Effectively
the Python interpreter checks input for syntax errors when using pyg4ometry classes. 
Another key benefit is the ability to use the pyg4ometry code to create 
programatic converters between different geometry languages.

>>>>

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{grey}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\lstset{
  showstringspaces=false,
  basicstyle=\ttfamily,
  columns=fullflexible,
  % frame=single,
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  language=Python,
  keywordstyle=\color{blue},
  breaklines=true,
  postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space},
}

@@

\title{PYG4OMETRY: a PYTHON library for the creation of Monte Carlo radiation transport physical geometries}

@@

\linenumbers

@@

\begin{small}
\noindent
{\em Program Title: PYG4OMETRY }                                         		\\
{\em Licensing provisions: GPLv3 }			\\
{\em Programming language: PYTHON, C/C++}                         		\\

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, radiation transport (RT) programmes, these 
include MCNP \cite{Mcnp_Werner}, FLUKA \cite{Fluka_Ferrari,Fluka_Bohlen}, Geant3 \cite{Geant3_Brun} and Geant4 \cite{Geant4_Agostinelli}. 
All these codes are based on the Monte Carlo technique but each code either has a particular specialism, simulation methodology or target user community.  
RT Monte Carlo (RTMC) simulations have diverse uses, shielding calculations for radiological protection, detector performance, medical imaging and therapy and space radiation environment  simulation are some examples. A fundamental requirement of all of the codes is to supply a computer-readable description of the 
physical geometry that  the particles are passing through.  The creation of geometry files is typically a very time consuming activity and the simulation validity 
and performance is directly dependent on the quality of the geometry. The requirement of a geometry system that allows the conversion between files prepared 
for different codes will enable cross-checks of the physics processes in different particle transport codes. In addition to the creation of geometry files for RT 
programs, usually computer assisted design (CAD) files exist for systems which need to be simulated. The fundamental geometric 
representations in CAD files is typically not amenable for RT programmes.  For these reasons it is advantageous to create a software tool that allows 
particle transport code users to rapidly develop error free geometry files, convert between common RT programme geometry formats and finally import CAD models. 
\colorbox{yellow}{More explanation of Geant4 and FLUKA}

This paper describes a geometry creation and conversion package called PYG4OMETRY, implemented in PYTHON and internally based on the Geant4 for APIs and Geometry Description Markup Language (GDML) for file persistency  \cite{GDML}. The main features of PYG4OMETRY are a PYTHON scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry descriptions, conversion from CAD formats  (STEP and IGES) based on FreeCAD \cite{FreeCAD} and OpenCascade \cite{OpenCASCADE} and powerful geometry visualisation tools based on VTK \cite{VTK4}. The origin of PYG4OMETRY was a set of utilities to prepare geometry for an accelerator beamline simulation based on Geant4 called BDSIM \cite{BDSIM_Nevay}. Accelerator physicists, like specialists in other areas, need a tool to quickly simulate beam losses and the subsequent interaction of the charged particle beam with the accelerator material. Figure~\ref{fig:workflow} depicts various workflows possible with PYG4OMETRY. PYG4OMETRY is not an executable software package but a toolkit, a user would typically write very small PYTHON program to use the PYG4OMETRY. This paper describes version 1.0 of PYG4OMETRY, which is freely available as a git repository and via PyPi.

\begin{figure*}[hbt!]
  \normalsize
  \centering
  \includegraphics[width=0.7\textwidth]{./diagrams/workflow.pdf}
  \caption{\label{fig:workflow}Schematic of PYG4OMETRY workflow, showing a) the different input file formats, b) PYTHON processing, c) output file targets and d) RTMC codes which use the geometry. }
\end{figure*}

There are existing codes that have functionality similar to PYG4OMETRY. Tools exist to convert CAD files to GDML, open source examples are GUIMesh \cite{GUIMesh_Pinto}, commercial products include ESABASE2 \cite{ESABASE2} and FASTRAD \cite{FASTRAD}. There are also tools from the fusion and 
neutronics community which  can convert CAD geometry into formats usable by RTMC codes, examples include DAGMC \cite{DAGMC} and McCad \cite{McCad}.  
CAD software can export data to STL. If a CAD model comprises of a large number of parts, exporting, assigning material to and placing the STL  components into 
the RTMC can be a cumbersome task. The existent set of software does not provide a complete set of tools to efficiently create complex geometries.

This paper is structured as follows, first a brief radiation transport-centric introduction to computer descriptions of geometry, followed by an explanation of the design and implementation of PYG4OMETRY. The sections that follow describe how PYG4OMETRY can be used to perform rapid geometric modelling, conversion from FLUKA to GDML, GDML to FLUKA and CAD to GDML. The paper concludes with an example of composite complex system consisting components of all the possible geometry input files formats.

\section{Computer descriptions of geometry} \label{sec:geometric}
Central to a computer-readable geometry is how a solid is defined in three dimensions. There are numerous different ways to describe a
geometry, including constructive solid geometry (CSG) , boundary representation (BREP) and tessellated polygons, these are described 
briefly in the following sections.  The Geant4 geometry specification is a mixture of geometry modelling techniques and described in detail last.

%%%%%%%%%%%%%%%%% Constructive solid geometry (CSG) 
Constructive solid geometry uses Boolean operations, subtraction, intersection and union, between simple solid shapes (e.g. cube, cylinder, sphere etc) or infinite
volumes (e.g a plane defined infinite half-space) to model complex surfaces which represent a solid. Boolean operations and solids can be combined to form a 
CSG tree to model complex geometry. FLUKA uses CSG to model solids, the form of Boolean expression used by FLUKA is not a general CSG tree but a 
logical expression in disjunctive normal form.

@@

%%%%%%%%%%%%%%%%% Tessellated polygons
Solid volumes can be defined using triangular, quadrilateral or tetrahedral meshes. Numerous formats exist to describe meshes, the ubiquitous being STL and 
more modern variants being PLY, OBJ to name a few. For solids with curved faces a tessellated mesh will always give an approximate description. As the mesh 
deviation distance from the solid decreases the number of polygons increases and hence the memory consumption and execution time of the RTMC simulation.

%%%%%%%%%%%%%%%%% Geant4 geometery
Geant4 geometry description is a mixture of BREP, CSG and tessellated concepts. Geant4 includes XX fundamental solids, although it does not store a sense 
of topology present in traditional CAD BREP systems. One of the fundamental solids is a {\em TessellatedSolid} which can used to represent STL or PLY files. 
Geant4 also provides the ability to perform Boolean operations on these primitive solids. The richest and most flexible geometry description is currently used by
Geant4. Not only do solid objects need to be defined but also placed in a world coordinate system. Geant4 has two concepts which facilitate this {\it Logical Volumes} and {\it Physical Volumes}. A {\it Logical Volume} is a region of space  that is defined by an outer solid but also other attributes like material, magnetic field 
and zero or more physical {\it daughter} volumes. A Physical Volume is a placement (or instance) of a Logical Volume.

%%%%%%%%%%%%%%%%% Geant Detector Mark-up Language
To exchange geometry descriptions between software packages Geometry Descripition Mark-up Language (GDML) was developed \cite{GDML}. 
GDML is an XML based description of Geant4 geometry description. Geant4 and ROOT \cite{fons_rademakers_2019_3895860} can read and write 
GDML format and it is commonly used as an exchange format for Geant4 geometries.

\section{PYG4OMETRY design and layout}
PYG4OMETRY is a PYTHON package consisting of semi-independent sub-packages. The sub-package \verb|pyg4ometry.geant4| contains all classes for 
geant4 detector construction, \verb|pyg4ometry.gdml| provides the functionality for reading and writing GDML files. There are sub-packages for importing and exporting other geometry formats  \verb|pyg4ometry.fluka|, \verb|pyg4ometry.stl| and \verb|pyg4ometery.freecad| and converting between file formats \verb|pyg4ometry.convert|.

The core of PYG4OMETRY consists of PYTHON classes that mimic Geant4 solids, logical, physical volume objects, GDML parameters and material classes.
The constructors of the PYTHON classes are kept as close to the original Geant4 C++ implementation so that PYG4OMETRY users do not have to learn 
a new API. For example the \verb|G4Box| class in Geant4, has the XML tag \verb|Box| in GDML and is represented by the \verb|geant4.solid.Box| class in 
PYTHON. Geometry construction in PYTHON proceeds in a way which is very similar to geometry construction in Geant4. A user relatively familiar with Geant4
should be able to start creating geometry in PYG4OMETRY immediately. In the following sections novel or important developments in PYG4OMETRY are described.

%\begin{table}[hbt!]
%\centering
%\begin{tabular}{ l l l } \hline
%GDML tag 		& Child tags				& PYG4OMETRY class 				\\ \hline
%Define 			& Constant				& gdml.Constant				\\
%				& Variable					& gdml.Variable					\\
%				& Quantity					& gdml.Quantity				\\
%				& Expression 				& gdml.Expression				\\
%				& Position					& gdml.Position					\\
%				& Rotation					& gdml.Rotation				\\
%				& Scale					& gdml.Scale					\\
%				& Matrix					& gdml.Matrix					\\ \hline
%Materials 		& Element					& geant4.Element				\\
%				& Isotope					& geant4.Isotope				\\
%				& Material					& geant4.Material				\\ \hline
%Solid  			& Box					& geant4.solid.Box				\\
%	  			& Tube					& geant4.solid.Tubs				\\
%	  			& CutTube				& geant4.solid.CutTubs			\\
%	  			& Cone					& geant4.solid.Cons				\\
%	  			& Para					& geant4.solid.Para				\\
%	  			& Trd					& geant4.solid.Trd				\\
%	  			& Trap					& geant4.solid.Trap				\\
%	  			& Sphere					& geant4.solid.Sphere			\\
%	  			& Orb					& geant4.solid.Orb				\\
%	  			& Torus					& geant4.solid.Torus				\\
%	  			& Polycone				& geant4.solid.Polycone			\\
%	  			& GenericPolycone			& geant4.solid.GenericPolycone	\\
%	  			& Polyhedra				& geant4.solid.Polyhedra			\\
%				& GenericPolyhedra                  & geant4.solid.GenericPolyhedra	\\ 
%	  			& Eltube					& geant4.solid.EllipticalTube		\\
%	  			& Ellipsoid					& geant4.solid.Ellipsoid			\\
%	  			& Elcone					& geant4.solid.EllipticalCone		\\
%				& Paraboloid				& geant4.solid.Paraboloid			\\
%				& Hype					& geant4.solid.Hyperboloid		\\
%				& Tet						& geant4.solid.Tet				\\
%				& Xtru					& geant4.solid.ExtrudedSolid		\\
%				& TwistedBox				& geant4.solid.TwistedBox		\\
%				& TwistedTtap				& geant4.solid.TwistedTrap		\\
%				& TwistedTrd				& geant4.solid.TwistedTrd 		\\
%				& TwistedTube				& geant4.solid.TwistedTube		\\
%				& Arb8					& geant4.solid.GenericTrap		\\
%				& Tessellated				& geant4.solid.TessellatedSolid 	\\
%				& Union					& geant4.solid.Union				\\
%				& Subtraction				& geant4.solid.Subtraction			\\
%				& Intersection				& geant4.solid.Intersection 		\\
%				& MultiUnion				& geant4.solid.MultiUnion 			\\
%				& ScaledSolid				& geant4.solid.ScaledSolid		\\				
%	  			& OpticalSurface			& geant4.solid.OpticalSurface		\\ \hline
%Structure		& Volume					& geant4.LogicalVolume			\\
%				& Assembly				& geant4.AssemblyVolume 		\\
%				& PhysVol					& geant4.PhysicalVolume			\\ 
%				& ReplicaVol				& geant4.ReplicaVolume			\\ 
%				& ParaVol					& geant4.ParameterisedVolume	\\ 
%				& DivisionVol				& geant4.DivisionVolume			\\
%				& SkinSurface				& geant4.SkinSurface			\\ 
%				& BorderSurface			& geant4.BorderSurface			\\ \hline
%
%\end{tabular}
%\label{tab:gdml-tags}
%\caption{GDML tags and corresponding PYG4OMETRY classes.}
%\end{table}

%%%%%%%%%%%%%%%%%  Input and output  
For each input format available in PYG4OMETRY  (GDML, STL, FLUKA and STEP) a dedicated {\it Reader} class is implemented, so \verb|gdml.Reader|, \verb|stl.Reader|, \verb|fluka.Reader| and \verb|freecad.Reader|. Each reader constructs the appropriate Geant4 classes and provides a \verb|Registry| which can be used or manipulated by the user. 
Output consists of taking the registry and writing the appropriate GDML.

%%%%%%%%%%%%%%%%%  Internal data representation
The internal data representation follows closely the structure of GDML. A \verb|Registry| class aggregates PYTHON ordered dictionaries that are  used to store the main 
elements of a GDML file. As a user specifies geometry the registry is updated. When a user is finished with the geometry, the registry is written to disk as a GDML file.
\verb|Registry| instances can be aggregated to form a composite geometry or volumes can be removed and added.

@@

\begin{lstlisting}[caption={A simple PYTHON script using PYG4OMETRY to create GDML variables},label={lst:gdmlExpressions}, language=Python]
# Import modules 
import pyg4ometry

@@

%%%%%%%%%%%%%%%%%  Replica, division, parametrised and looped volumes
A powerful feature of Geant4 and hence GDML is the ability to either repeat, divide or parametrise geometry. The create multiple replicas of a volume in a Cartesian, cylindrical or spherical grid is known as a ReplicaVolume. A DivisionVolume breaks a primitive into segments in either a cartesian or cylindrical polar coordinates. 
A Parametrised Volume allows for the arbitrary multiple placement of solids where the parameters are allowed to vary for each placement.  GDML supports loops, 
where sections of GDML are allowed to be repeated with varying parameters. GDML loops are not supported by PYG4OMETRY and will be implemented in a 
future release.

\subsection{Tessellation of solids (meshing)}
For each Geant4 primitive solid a triangular tessellated Vertex-Face mesh is generated and cached. This mesh is then used to determine the extent of placed 
instances of geometry (physical volumes) and meshes for CSG derived solids. CSG mesh calculations are performed using a BSP tree technique in pure PYTHON 
\cite{pycsg} or the Computational Geometry Algorithms Library (CGAL) surface meshes \cite{cgal:bsmf-sm-20b}. In general the CGAL implementation is one or two orders of magnitude faster than the pure-PYTHON CSG implementation and must be used for large geometries. Triangular meshes based on CSG 
operations involving curved surfaces often contain large numbers of thin triangles. Before meshes are visualised or written to disk various polygon mesh 
algorithms  from CGAL \cite{cgal:lty-pmp-20b} can be employed to give the meshes more desirable features.

@@

The {\it Physical Volume} class (\verb|pyg4ometry.geant4.PhysicalVolume|) is also used to store visualisation attributes like the solid's 
colour, surface or wireframe representation and visibility. Overlaps detected in the mesh geometry are
stored in the Logical Volume and can be  displayed separately to allow a user to understand the reason for 
the overlaps.

%%%%%%%%%%%%%%%%% Adding properties to volumes
Geometry needs to be augmented with other information for a complete RTMC simulation. Often other attributes need to be 
assigned to regions of space for example material definition, magnetic field or optical properties. These physical properties 
can be used to define the visualisation attributes of a volume.

\subsection{Overlap detection}
All RTMC codes can not handle spatial overlap between two geometric objects and likely are to have ill-defined behaviour when tracking particles  
through such a situation.  A key feature of PYG4OMETRY is the detection of potential overlaps efficiently, it does this by performing an intersection operation between solid instances and determining if the resulting mesh is empty. Figure~\ref{fig:overlap} shows the potential overlaps, a) protrusion of a daughter from the mother, b)  finite volume intersection between two daughters  and c) an overlap where two daughters share a face. 
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{./diagrams/overlap.pdf}
\caption{Schematic of the three different types of overlap between daughters of a mother logical volume.}
\label{fig:overlap}
\end{center}
\end{figure}

@@

Overlap detection in PYG4OMETRY relies of the meshes generated for each solid, 
any detection algorithm will be approximate. Generally the overlap detection 
algorithm proceeds as 
\begin{enumerate}
\item Check if any of the daughter solid meshes have a non-null intersection with the compliment of mother solid mesh
\item Loop over all daughter volumes and check
\item Check
\end{enumerate}

@@

%%%%%%%%%%%%%%%%% Graphical user interface
% A basic Qt based graphical user interface is also available with PYG4OMETRY. If a model has a large number of volumes defined in a hierarchical tree structure, it is % cumbersome for the user to find the logical or physical volume.

\section{Rapid geometry modelling}
Given the PYTHON scripting interface, expression  and tessellation engines. It is possible for a user to rapidly specify the geometrical layout of the RT problem, vary 
the parameters of the geometry and visualise it.  When a user has achieved the desired geometry, without geometry overlaps, a GDML file can be written to disk from the internal memory representation. Some example code is presented in Listing~\ref{lst:pythonRapidModelling}. The structure should be familiar to regular 
users of Geant4 or GDML, apart from a new class called the \verb|Registry|. First a \verb|Registry| need to be created to store all the PYG4OMETRY objects; followed by constants;  then materials and properties, solids and logical and physical volumes; finally the whole geomtery can be saved as a GDML file and/or visualised using VTK.

\begin{lstlisting}[caption={A simple PYTHON script using PYG4OMETRY to create a simple Geant4 geometry.},label={lst:pythonRapidModelling}, language=Python]
# Import modules 
import pyg4ometry.gdml as gd
import pyg4ometry.geant4 as g4
import pyg4ometry.visualisation as vi

@@

The PYG4OMETRY PYTHON code in the example is approximately as expressive 
as the GDML it writes. The benefit of wrapping GDML/Geant4 in PYTHON allows 
to very rapid prototyping of geometry without concerns of compilation of C++ (in the 
case of Geant4) or writing well formed XML (in the case of GDML). Effectively
the PYTHON interpreter checks input for syntax errors when using PYG4OMETRY classes. 
Another key benefit is the ability to use the PYG4OMETRY code to create 
programatic converters between different geometry languages or more generally 
manipulation and transformation of the geometry stored in memory. The rapid modelling 
example given in Listing~\ref{lst:pythonRapidModelling} and Figure~\ref{fig:rapidModellingExample} 
is rather trivial, a significantly more complex example is in Figure~\ref{fig:gdml-flair}.
